---
title: "STA 141 - Fundamentals of Statistical Data Science - Final project"
params:
  term: Winter 2023
  duedate: '2023-03-24'
  name: ERIC SUN + WILSON ZHOU
  email: etsun@ucdavis.edu
  total_points: XX
  grade: NA
output: 
   pdf_document: default
   html_document: default
---




# Working with the New York Times COVID-19 Data

As we have seen in class, COVID-19 data collected by the *New York
Times* are available in a repository on Github.  In this final project we
will work with these data to reproduce some of the elements of the
*Times*'s reporting on COVID-19.  

```{r include = FALSE, echo = TRUE} 
diff.eg <- quote(diff(c(1, 3, 6, 10)))
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
```


# Exercises

1. The file `us.csv` contains aggregated data for the entire U.S.  In
   this file, the `cases` and `deaths` columns represent *cumulative*
   cases and deaths due to covid. The `diff()` function can be used to
   compute the differences between each consecutive element of a
   vector, so it could be used to compute the daily numbers of cases
   and deaths.  However, `diff()` returns a vector of length one less
   than the length of the original vector (e.g., `r noquote(deparse(diff.eg))`
   returns `r eval(diff.eg)`) and this can make it somewhat
   inconvenient to use when transforming columns of data frames.
   
   An alternative is to us the more general `filter()` function with
   an appropriate choice of the `filter` and `sides` arguments.  Thus function
   can also be used to computing running averages and similar quantities.
   
    a. Read the file `us.csv` into R as the data frame `us` and do the
       following:
       
```{r}
us <- read.csv("us.csv")
# transform to Date class
us$date <- as.Date(us$date)
```
          
```{r}
us$cases = as.ts(us$cases)
us$new_cases <- c(stats::filter(us$cases, c(1,-1), sides=1))
```
          
```{r}
us$deaths = as.ts(us$deaths)
us$new_deaths <- c(stats::filter(us$deaths, c(1,-1), sides=1))
```
     	    
```{r}
us$avg_new_cases = c(stats::filter(us$new_cases, rep.int(1/7,7), sides = 1))
us$avg_new_cases = round(us$avg_new_cases,2)
```
     	    
```{r}
us$avg_new_deaths = c(stats::filter(us$new_deaths, rep.int(1/7,7), sides = 1))
us$avg_new_deaths = round(us$avg_new_deaths,2)
```
			
      
    b. Create a plot of daily cases similar to the one found at the top of [this 
       page](https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html).
       Plot only data beginning from 2020-03-01.
       
```{r, echo = FALSE}
us_new = subset(us, date >= "2020-03-01")

plot(us_new$date, us_new$new_cases, type = "h", col = "salmon",
     main = "Daily Cases of COVID19 in the US from March 2020-",
     xlab = "Dates",
     ylab = "Cases")
lines(us_new$date, us_new$avg_new_cases, col = "red", lwd = 2)
legend(x = "topright",legend = c("Daily Cases","Average New Cases"), col = c("salmon","red"),  
       lwd = 2) 
```

    c. Repeat part (b) for deaths.
    
```{r, echo = FALSE}
plot(us_new$date, us_new$new_deaths, type = "h", col = "salmon",
     main = "Daily Deaths of COVID19 in the US from March 2020-",
     xlab = "Dates",
     ylab = "Deaths")
lines(us_new$date, us_new$avg_new_deaths, col = "red", lwd = 2)
legend(x = "topright",legend = c("Daily Deaths","Average New Deaths"), col = c("salmon","red"),  
       lwd = 2) 
```
    

2. The file `us-states.csv` contains county-level data for the U.S.

    a. Read `us-states.csv` into R as the data frame `us_states`
       and transform the date column into a column of class `Date`.
	   
```{r}
us_states <- read.csv("us-states.csv")
us_states$date <- as.Date(us_states$date)
```


    b. Use `subset()` to extract the data for the state of California
       and save it as a data frame named `California`. Be sure that the
       rows are correctly ordered by date, and then repeat parts 1b
       and 1c of this assignment for California, i.e., plot the number of
       daily new cases and deaths, along with their 7-day running averages.

```{r, echo = FALSE}
California = subset(us_states, state == "California")
California = California[order(California$date),]

California$new_cases <- c(stats::filter(California$cases, c(1,-1), sides=1))
California$new_deaths <- c(stats::filter(California$deaths, c(1,-1), sides=1))
California$avg_new_cases = c(stats::filter(California$new_cases, rep.int(1/7,7), sides = 1))
California$avg_new_cases = round(California$avg_new_cases,2)
California$avg_new_deaths = c(stats::filter(California$new_deaths, rep.int(1/7,7), sides = 1))
California$avg_new_deaths = round(California$avg_new_deaths,2)

California_new = subset(California, date >= "2020-03-01")

plot(California_new$date, California_new$new_cases, type = "h", col = "salmon",
     main = "Daily Cases of COVID19 in California from March 2020-",
     xlab = "Dates",
     ylab = "Cases")
lines(California_new$date, California_new$avg_new_cases, col = "red", lwd = 2)
legend(x = "topright",legend = c("Daily Cases","Average New Cases"), col = c("salmon","red"),  
       lwd = 2)  

plot(California_new$date, California_new$new_deaths, type = "h", col = "salmon",
     main = "Daily Deaths of COVID19 in California from March 2020-",
     xlab = "Dates",
     ylab = "Deaths")
lines(California_new$date, California_new$avg_new_deaths, col = "red", lwd = 2)
legend(x = "topright",legend = c("Daily Deaths","Average New Deaths"), col = c("salmon","red"),  
       lwd = 2) 
```


3. The file `us-counties.csv` contains county-level data for the U.S.

    a. Read `us-counties.csv` into R as the data frame `us_counties`
       and transform the date column into a column of class `Date`.

```{r}
us_counties <- read.csv("us-counties.csv")
us_counties$date <- as.Date(us_counties$date)
```

    b. Use `subset()` to extract the data for Yolo County, California,
       and save it as a data frame named `Yolo`. Be sure that the
       rows are correctly ordered by date, and then repeat part 1b
       this assignment for Yolo County, i.e., plot the number of daily new
       cases along with their 7-day running average. 
	   
```{r, echo = FALSE}
Yolo = subset(us_counties, county == "Yolo")
Yolo = Yolo[order(Yolo$date),]

Yolo$new_cases <- c(stats::filter(Yolo$cases, c(1,-1), sides=1))
Yolo$new_deaths <- c(stats::filter(Yolo$deaths, c(1,-1), sides=1))
Yolo$avg_new_cases = c(stats::filter(Yolo$new_cases, rep.int(1/7,7), sides = 1))
Yolo$avg_new_cases = round(Yolo$avg_new_cases,2)
Yolo$avg_new_deaths = c(stats::filter(Yolo$new_deaths, rep.int(1/7,7), sides = 1))
Yolo$avg_new_deaths = round(Yolo$avg_new_deaths,2)

Yolo_new = subset(Yolo, date >= "2020-03-01")

plot(Yolo_new$date, Yolo$new_cases, type = "h", col = "salmon",
     main = "Daily Cases of COVID19 in Yolo County from March 2020-",
     xlab = "Dates",
     ylab = "Cases")
lines(Yolo_new$date, Yolo_new$avg_new_cases, col = "red", lwd = 2)
legend(x = "topright",legend = c("Daily Cases","Average New Cases"), col = c("salmon","red"),  
       lwd = 2) 

plot(Yolo_new$date, Yolo_new$new_deaths, type = "h", col = "salmon",
     main = "Daily Deaths of COVID19 in Yolo County from March 2020-",
     xlab = "Dates",
     ylab = "Deaths")
lines(Yolo_new$date, Yolo_new$avg_new_deaths, col = "red", lwd = 2)
legend(x = "topright",legend = c("Daily Deaths","Average New Deaths"), col = c("salmon","red"),  
       lwd = 2) 
```
	   
	   
	   Q: What do you notice when comparing the plot of daily new cases
     in Yolo county to the analogous plot for the state of
	   California as a whole?  What might explain what you are seeing?
	   
	   <Br>
	   
	   The plots of daily new cases for both Yolo County and California seem to have similar trends/distributions.
	   They both had distinct peaks, indicating a rapid influx of cases, around 2021 and especially during the start
	   of 2022. 
	   
	   (Note: data for Yolo County has not been updated since May 2022, so
	   we cannot visualize how the trends compare to California beyond then). 
	   
	   However, when we take the peak in the year 2022, the cases in Yolo county compared to the cases in California as a        whole accounts for 0.5% of the total in California. This could be the case because the people of Yolo county are very      careful and good at practicing social distancing.
	   The reason for these peaks may be due to a variety of reasons, including: origin of new variants,
	   looser restrictions on mask + social distancing policies and enforcement,
	   increased travel/vacation plans during winter break that led to greater risk 
	   of obtaining the virus, etc.
	   
	   Additional Note: Some graphs display negative values for Daily Cases and Daily Deaths.
	   This indicates an error in the data collection process, as it is not possible to have
	   a case or death count below 0.
	 
### __Negative Values__
```{r}
us_new[us_new$new_cases < 0,] # US data: new cases
us_new[us_new$new_deaths < 0,] # US data: new deaths
California_new[California_new$new_cases < 0,] # Cali data: new cases
California_new[California_new$new_deaths < 0,] # Cali data: new deaths
Yolo_new[Yolo_new$new_cases < 0, ] # Yolo data: new cases
Yolo_new[Yolo_new$new_deaths < 0, ] # Yolo data: new deaths
```

4. Bonus 	   

```{r, echo = FALSE}
library(maps) # gives us pre-drawn map data

states <- map_data("state") # map data for states

us_states = us_states %>% arrange(state)

us_states$new_cases <- c(stats::filter(us_states$cases, c(1,-1), sides=1))
us_states$new_deaths <- c(stats::filter(us_states$deaths, c(1,-1), sides=1))

us_states$avg_new_cases = c(stats::filter(us_states$new_cases, rep.int(1/7,7), sides = 1))
us_states$avg_new_cases = as.integer(us_states$avg_new_cases)

us_states$avg_new_deaths = c(stats::filter(us_states$new_deaths, rep.int(1/7,7), sides = 1))
us_states$avg_new_deaths = round(us_states$avg_new_deaths,2)
  
us_states$region = tolower(us_states$state)
us_states_heatmap = left_join(states, us_states, multiple = "all")

us_states_heatmap = subset(us_states_heatmap, date > "2023-03-13")
us_states_heatmap
```

### __Heatmaps for States: COVID-19 Avg Cases and Deaths__
```{r, echo = FALSE}
ggplot(data = us_states_heatmap, mapping = aes(x = long, y = lat, group = group, fill = avg_new_cases)) +
  geom_polygon(color = "gray90", size = .05) + coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme(legend.position = "bottom") + labs(title = "Average New COVID-19 Cases per State", fill = "Average New 
                                           Cases per week")

ggplot(data = us_states_heatmap, mapping = aes(x = long, y = lat, group = group, fill = avg_new_deaths)) +
  geom_polygon(color = "gray90", size = .05) + coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme(legend.position = "bottom") + labs(title = "Average New COVID-19 Deaths per State", fill = "Average New Deaths per week")
```


```{r, echo = FALSE}
counties <- map_data("county")

us_counties = us_counties %>% arrange(county)

us_counties$new_cases <- c(stats::filter(us_counties$cases, c(1,-1), sides=1))
us_counties$new_deaths <- c(stats::filter(us_counties$deaths, c(1,-1), sides=1))
us_counties$avg_new_cases = c(stats::filter(us_counties$new_cases, rep.int(1/7,7), sides = 1))
us_counties$avg_new_cases = as.integer(us_counties$avg_new_cases)
us_counties$avg_new_deaths = c(stats::filter(us_counties$new_deaths, rep.int(1/7,7), sides = 1))
us_counties$avg_new_deaths = round(us_counties$avg_new_deaths,2)

us_counties$subregion = tolower(us_counties$county)
us_counties <- us_counties %>% arrange(state,subregion)
us_counties = subset(us_counties, date > "2022-05-06")

us_counties_heatmap = left_join(counties, us_counties, by = "subregion", multiple = "all")
```

### __Heatmaps for Counties for Last Week in Dataset (Week of May 7th, 2022): COVID-19 Avg Cases and Deaths__
```{r, echo = FALSE}
ggplot(data = us_counties_heatmap, mapping = aes(x = long, y = lat, group = group, fill = avg_new_cases)) +
  geom_polygon(color = "gray90", size = .05) + coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme(legend.position = "bottom") + labs(title = "Average New COVID-19 Cases per County", fill = "Average New Cases per week")

ggplot(data = us_counties_heatmap, mapping = aes(x = long, y = lat, group = group, fill = avg_new_deaths)) +
  geom_polygon(color = "gray90", size = .05) + coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme(legend.position = "bottom") + labs(title = "Average New COVID-19 Deaths per County", fill = "Average New Deaths per week")
```


# Code Appendix
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
