vf---
title: "STA141A HW1"
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Macbook Specs:
- 16 inch (2019)
- 8-Core Intel i9
- 16 GB
- Intel UHD Graphics 630 1536 MB

#Problem 1 - Defining Function Coupon
```{r}
coupon_mean_sd <- function(n) {
    p <- 1-(0:(n-1))/n
    m <- sum(1/p)
    sd <- sqrt(sum((1-p)/(p^2)))
    return(c(m,sd))
}
```

```{r}
coupon_mean_sd(25)
coupon_mean_sd(50)
coupon_mean_sd(100)
```

#Problem 2 - Comparing slow and fast for different replications and sample sizes
```{r}
#slow
coupon_slow <- function(n) {
  coupons <- rep(0L, n)
  while (any(coupons == 0L)) {
    new_coupon <- sample.int(n, 1)
    coupons[new_coupon] <- coupons[new_coupon] + 1
    }
  sum(coupons)
}

#fast
coupon_fast <- function(n) {
  p <- 1 - (0:(n-1))/n
  x <- rgeom(n, p) ## R counts only failures
  n + sum(x) ## so add n * 1 = n to get the total
}
```

```{r}
system.time(replicate(10000,coupon_slow(25)))
system.time(replicate(10000,coupon_fast(25)))

system.time(replicate(10000,coupon_slow(50)))
system.time(replicate(10000,coupon_fast(50)))

system.time(replicate(10000,coupon_slow(100)))
system.time(replicate(10000,coupon_fast(100)))
```
The ratio at which coupon_slow is slower than coupon_fast depends on the number
of trials. The greater the number of trials, the greater the ratio. 

#Problem 3 
#a.)
```{r}
S <- replicate(10000,coupon_fast(50))
mean(S)
sd(S)
```
The sample and theoretical mean and sd have a small marginal error, so they
agree reasonably well. 

#b.) 
```{r}
hist(S, nclass = 25)
```

#c.)
```{r}
quantile(S, probs=c(.1,.5,.9))
```

#Problem 4
```{r}
mcoupon_fast <- function(n, m) {
  coupons <- rep(0L, n) # Will count each one until we have m of that coupon, then stop, print out n 0s
  incomplete <- (coupons < m) #returns TRUE if there is a value in coupon less than m
  S <- 0 #initial coupon count
  while (any(incomplete)) { #if condition for m is not yet satisfied
    k <- sum(incomplete) # Number of coupons with fewer than m accumulated so far
    x <- rgeom(1, k/n) + 1 # Successes getting any one of the coupons we still need, rgeom returns number of failures before success
    S <- S + x # Update number of purchases
    y <- sample.int(k, 1) # Which of the not-yet-completed coupons did we draw
    coupons[incomplete][y] <- coupons[incomplete][y] + 1 # Increment incomplete coupon counts
    incomplete <- (coupons < m)
    }
  S
}
```

#Mcoupon_slow Function
```{r}
mcoupon_slow <- function(n,m) {
  coupons <- rep(0L, n)
  incomplete <- (coupons < m)
  while (any(incomplete)) {
    new_coupon <- sample.int(n, 1)
    coupons[new_coupon] <- coupons[new_coupon] + 1
    incomplete <- (coupons < m)
    }
  sum(coupons)
}
```

#b.)
```{r}
m2 <- replicate(10000,mcoupon_fast(50,2))
mean(m2)
sd(m2)
quantile(m2, probs=c(.1,.5,.9))
```

#c.)
```{r}
m3 <- replicate(10000,mcoupon_fast(50,3))
mean(m3)
sd(m3)
quantile(m3, probs=c(.1,.5,.9))
```

#Bonus

```{r}
head(mtcars)
```
#Regression Model
```{r}
reg <- lm(mpg ~ wt + disp, data = mtcars)
```

#Question 1
```{r}
X <- model.matrix(reg) #design matrix
Y = mtcars$mpg

QR <- qr(X)
Q <- qr.Q(QR)
R <- qr.R(QR)
```

#Solutions to XTX
```{r}
x <- solve(t(X)%*%X)
x1 <- solve(t(Q%*%R)%*%(Q%*%R))

x
x1
```

#Question 2
```{r}
standard_res <- sqrt(abs(reg$residuals/sd(reg$residuals)))
plot(reg$fitted.values, standard_res, main = "Scale-Location Plot",
     xlab = "Fitted Values", ylab = "Root of Standard Residuals",
     ylim = c(0,1.5))
lines(lowess(reg$fitted.values,standard_res), col = 'red')
```
Extra Notes:
Scale-location plot is a plot that displays
fitted values of regression model along x-axis and
square root of standardized residuals along y-axis

Check that:
- red line is roughly horizontal (homoscedastic)
- no obvious patterns; means randomly scattered errors; roughly equal variability at all fitted values

# Code Appendix
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
