---
title: "STA 141A Fundamentals of Statistical Data Science - Homework 2"
params:
  term: Winter 2023
  duedate: '2023-03-08'
  name: ERIC SUN
  email: ETSUN@UCDAVIS.EDU
  total_points: XX
  grade: NA
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# packages
library(ggplot2)
```


```{r as_ordinal, include = TRUE, echo = FALSE}
ord_suffix <-
    function(x) switch(as.character(x), "1"="st", "2"="nd", "3"="rd", "th")
as_ordinal <- function(x) paste0(x, ord_suffix(x))
```

# Exercises

1.After executing the command

```{r eval=FALSE, include=TRUE}
z <- rep(c(rep(x, each = a), rep(y, each = b)), times = t)
```
which element of x or y is contained in the kth element of z?. Suppose that x and y are vectors in R of lengths m and n,
respectively.  Write a function, `rep_tracker(k, m, n, a, b, t)`, which computes
the answer for any positive integers k, m, n, a, b, and t.  Your
function should return a list whose first element is named
"src_vector" and gives the name of the vector ("x" or "y") and
whose second element is named "src_index" and gives the integer
indicating the original location of the value in either "x" or "y".
For example, if the kth element of z was originally the 5 element
of y, then your function should return `list(src_vector = "y",
src_index = 5)`.

Your may not use the `rep()` function in your solution.  Your function
should emulate the way that a human might solve this problem using
integer arithmetic.

# Solution

```{r, echo = TRUE}
rep_tracker <- function(k, m, n, a, b, t) {
    if (k > (t * (m * a + n * b))) warning("First argument is out of range")
    
    r <- k %% (m*a + n*b) #remainder
    
    if (r == 0L){ # if remainder = 0; it is in y
      v = "y"
      i = n # index is just last value of y (length of y)
    }
    else if (r <= m*a){
      v = "x"
      d = a
      i = ceiling(r/d) # use remainder divided by length of x plus (ceiling: to account for repetitions) to find index 
    } else {
      v = "y"
      r = r - (m*a)
      d = b
      i = ceiling(r/b)
    }
    
    list(src_vector = v, src_index = i)
}
```

<Br>
<Br>

##### __2. Consider the report:  [A PHASE 1/2/3, PLACEBO-CONTROLLED, RANDOMIZED, OBSERVER-BLIND, DOSE-FINDING STUDY TO EVALUATE THE SAFETY, TOLERABILITY, IMMUNOGENICITY, AND EFFICACY OF SARS-COV-2 RNA VACCINE CANDIDATES AGAINST COVID-19 IN HEALTHY INDIVIDUALS (pages 99-101)]. Let $\theta$ be the probability that a subject who fell ill with Covid-19 is from the treatment group and $1-\theta$  the probability that the subject is from the control group. Assuming that 94 subjects fell ill to Covid-19 (with a sample efficacy above 90%) and at most 8 of those 94 subjects were vaccinated. Write a report (Introduction, Methods, Results and Conclusions) assuming:__

<Br>

 __a. A Beta prior for $\theta$: $p(\theta)$=Beta$($a=0.5,b=0.5$)$, where $a$ and $b$ are the shape parameters
 of the Beta distribution. Plot the prior, likelihood and posterior as function of $\theta$.__
 
<Br>

#### __Introduction__

To test the effectiveness and safety of their new vaccine against COVID-19, Biontech and Pfizer conducted a case study using the Bayesian 
analysis framework to update and uphold their credible beliefs on the matter. Given that the two groups (treatment vs control) were assigned the same number of subjects, $\theta$ represented the probability that a subject who fell ill with COVID-19 was from the treatment and $1-\theta$ represented the probability for the control group. 
In their study, they assumed having a prior distribution of:
__$$\theta=(1−VE)/(2−VE)$$__
The assumed prior distribution of $\theta$ for this study is a Beta distribution with shape parameters a=a0=0.5 and b=b0=0.5. 
The prior mean of $\theta$ is thus given by:
__$$E(\theta)=a_{0}/(a_{0}+b_{0})=0.5$$__
Therefore, the vaccine efficacy:
$$VE=(1−2\theta)/(1−\theta)$$
at $\theta=0.5$ is 0%, as proposed by Biontech and Pfizer. This makes sense, as our prior parameters reflect that there is an equal probability of getting COVID-19 whether a subject is vaccinated or not, which questions the true effectiveness of the vaccine. In addition for this study, we have 94 ill subjects with a sample efficacy above 90%, and we assume that at most 8 of those subjects that had COVID-19 were vaccinated. To test whether the vaccine is effective, I will construct the prior, likelihood, and posterior models, evaluate the credible, vaccine efficacy, and confidence intervals, and plot the posterior empirical predictive density model.


<Br>

#### __Models__

```{r, echo = FALSE}
a = 0.5
b = 0.5
n = 94 #number of ill subjects
successes = 8 #treatment group
theta = seq(0, 1, length = 1000)

#likelihood and posterior parameters
likelihood_a = successes + 1; likelihood_b = n - successes + 1
posterior_a = a + successes; posterior_b = b + n - successes

#prior,likelihood, and posterior distributions
prior <- dbeta(theta,a,b)
likelihood <- dbeta(theta, likelihood_a, likelihood_b)
posterior <- dbeta(theta, posterior_a, posterior_b)

a0 = 0.5; b0 = 0.5
ggplot(data = data.frame(x = c(0, 1)), aes(x)) +
  stat_function(geom="area",fun = dbeta, n = 101,
    args = list(shape1 = a0, shape2 = b0),
    col="red", fill="red", alpha=0.5
  ) +
  ylab("Prior") + xlab(expression(theta)) + geom_vline(xintercept=0.5)+
  ggtitle("Prior: Probability that a subject with Covid-19 was vaccinated")  

a1 = likelihood_a; b1 = likelihood_b
ggplot(data = data.frame(x = c(0, 1)), aes(x)) +
  stat_function(geom="area",fun = dbeta, n = 101,
    args = list(shape1 = a1, shape2 = b1),
    col="blue", fill="blue", alpha=0.5
  ) +
  ylab("Likelihood") + xlab(expression(theta)) + geom_vline(xintercept = a1/(a1+b1))+
  ggtitle("Likelihood that a subject has COVID-19 given theta") 

a2 = posterior_a; b2 = posterior_b
ggplot(data = data.frame(x = c(0, 1)), aes(x)) +
  stat_function(geom="area",fun = dbeta, n = 101,
    args = list(shape1 = a2, shape2 = b2),
    col="black", fill="red", alpha=0.5
  ) +
  ylab("Posterior") + xlab(expression(theta)) + geom_vline(xintercept=a2/(a2+b2))+
  ggtitle("Posterior if 8 out of 94 subjects were vaccinated")  
```
<Br>

<Br>

#### __Interpretation__

Our parameters for the prior (a = b = 0.5) generated a U-shaped, symmetrical graph
with the center at $\theta=0.5$. For the likelihood (parameters = (9,87)) and posterior (parameters = (8.5,86.5)) graphs, the distribution 
and shape are almost identical. This is due to the prior being non-informative, which 
means that theta is mainly determined by the data. Because of this, the posterior shifted
far down to the left, which decreased our theta based on the observations. The posterior
shows that our possible distribution for the theta parameter distribution is lower than what is was
in the prior, and indicates that the probability of a vaccinated subject that has COVID-19
is actually smaller than the probability of a subject having COVID-19 in the control group.


<Br>

 __b. Compute the posterior probability of having a value of $\theta>0.4118$.__
 
```{r, echo = FALSE}
prob = 1 - pbeta(0.4118,posterior_a,posterior_b)
cat("The posterior probability of having a value of theta greater than .4118 is",prob)
```

Biontech/Pfizer stated as interim success criterion that the posterior probability of an efficacy below 30% (corresponding to $\theta>0.4118$) is smaller than 2.5%. After conducting the posterior, we conclude that the probability that a subject that was
vaccinated has COVID-19 is 5.82645e-13 at $\theta>0.4118$, which is almost 0%. We see that the probability of observing the posterior probability of an efficacy below 30% is extraordinarily small, thus passes the interim success criterion. We can also state that since the probability of the probability that a subject who fell ill is from the treatment group that is greater than theta > .4118 is very small, it indicates that those in the treatment group have a greater chance of not having COVID-19 compared to those in the control, which shows the vaccines effectiveness.

<Br>

 __c. Compute a 95\%  credible and confidence intervals.__
 
```{r, echo = FALSE}
#credible interval
theta.credible.interval = qbeta(c(0.025,0.975),posterior_a,posterior_b)
cat("The values for the credible interval are",round(theta.credible.interval*100,1))

#VE interval
VE.credible.interval = rev((1-2*theta.credible.interval)/(1-theta.credible.interval))
cat("\nThe values for the vaccine efficacy interval are",round(VE.credible.interval*100,1))

#confidence interval
confidence = qbeta(c(.025,.975),successes+1,n-successes+1)
cat("\nThe values for the confidence interval are",round(confidence*100,1))

VE.conf.interval = rev((1-2*confidence)/(1-confidence))
cat("\nThe values for the vaccine efficacy confidence interval are",round(VE.conf.interval*100,1))
```

Given our belief and data, we believe with 95% probability that the probability
that an ill subject is vaccinated is between 4.1% and 15.4%. This means that
we would be 95% confident that the vaccine reduces the risk of getting COVID
between 81.8% and 95.7%. For the confidence interval, we are 95% confident that
an ill subject is vaccinated is between 4.4% and 15.9%. This can also be translated as
we are 95% confident that the vaccine reduces the risk of getting COVID between
81.1% and 95.4%. We observe that the Bayesian credible interval is close to the
frequentist confidence interval.


<Br>

 __d.  Plot the posterior empirical predictive density. If a new sample of 94 subjects with Covid-19 is taken, how many subjects were vaccinated?.__

```{r, echo = FALSE}
#generate 10k simulations of posterior distribution
post_data = rbeta(10000,posterior_a,posterior_b)
#histogram for theta vales
hist(post_data, main = "Theta Distribution for Posterior", xlab = "Thetas")
abline(v = posterior_a/(posterior_a+posterior_b), col = "red")

#posterior predictive density
post_density <- rbinom(10000,94,post_data)
hist(post_density, freq = FALSE, main = "Posterior Emperirical Predictive Density",
     xlab = "# of Subjects")
abline(v = posterior_a/(posterior_a+posterior_b) * 94, col = "red")

#expected number of subjects vaccinated for a new sample
expected <- posterior_a/(posterior_a+posterior_b)
subjects <- expected*94
cat("The expected amount of vaccinated subjects if we have a new sample of 94 subjects is:", subjects)
```

I generated 10000 samples from the posterior distribution to obtain a range
of practical thetas (post_data). We see that most of our data lies
within the .05-.15 range, and the mean, median, and standard deviation are .08932,
.08687, and .02883, respectively. After generating the distribution of thetas, I 
generated the posterior predictive density plot by plugging in the post_data into
the parameters for rbinom(). To calculate how many subjects were vaccinated for
a new sample of 94 subjects, I used the previous definition of the expected theta, 
plugged in the posterior parameters, and multiplied the expected theta by the number
of subjects. The result was 8.410526 subjects ~> 8 subjects.

<Br>

#### __Conclusion__

From our results, we saw that:

1.) The prior, likelihood, and posterior plots show how the theta distribution is more condensed towards smaller values of theta (posterior), which indicates that the probability of a vaccinated subject having COVID-19 is smaller compared to the control group.

2.) The criteria for success at an interim analysis based on the posterior probability (P[VE > 30% given data]) at the current number of cases was met. 

3.) The confidence and credible intervals show that around ~ min[4] to max[16] %  of vaccinated people are ill with 95% confidence. The VE intervals show that having the vaccine reduces the risk of catching COVID by ~ min[80] to max[96]%.

Thus, we can conclude that subjects who are vaccinated have a much greater chance of not getting COVID-19 compared to those who are not.

<Br> 
<Br>


# Code Appendix
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```






